Size-differentiated Export in different Dynamical Regimes in the Ocean

### 2019-02-14 size_differentiated_export.tex
Updates from Overleaf

%% ------------------------------------------------------------------------ %%
%
%  AUTHORS AND AFFILIATIONS
%
%% ------------------------------------------------------------------------ %%
\authors{M. Dever\affil{1}, D. Nicholson\affil{1}, M. Omand\affil{2}, and A. Mahadevan\affil{1}}
\affiliation{1}{Woods Hole Oceanographic Institution, Woods Hole, MA 02543, USA}
\affiliation{2}{Graduate School of Oceanography, University of Rhode Island, Narragansett, Rhode Island, USA}

@@

%% ------------------------------------------------------------------------ %%
%
%  ABSTRACT
%
%% ------------------------------------------------------------------------ %%
\begin{abstract}
% Mat: I think that even with the submesoscale mechanism, export is mainly driven by gravitational sinking. So I'm moving the "Traditionally" to later. 
Export of particulate organic carbon (POC) is  mainly driven by gravitational sinking. Thus, traditionally, it is thought that  larger, faster-sinking particles make up most of the POC export flux. However, this need not be the case in a dynamic oceanic flow field, where the ocean velocity can influence the descent rate of particles. Particles with different settling speeds are released in two process-oriented model simulations of an upper ocean eddying flow to evaluate the impact of (1) the particle number size-spectrum slope, and (2) the ocean dynamics on the respective contribution of the different sinking-velocity classes to POC export. The analysis reveals that the leading export mechanism changes from gravitationally-driven to advectively-driven as submesoscale dynamics become more important. The vertical velocity associated with submesoscale dynamics enhances the contribution of slower-sinking particles in POC export. A steeper particle size spectrum, also increases the contribution of smaller, slower-sinking particles. Implementing a remineralization scheme generally decreases the total amount of biomass exported, but its impact is weaker in dynamical regimes where submesoscale dynamics are present and export is advectively-driven. Under specific conditions, remineralization processes counter-intuitively enhance the role of slower-sinking particles to the point where these slower-sinking velocity classes dominate the export, therefore challenging the traditional paradigm for POC export. This study demonstrates that slow-sinking particles are significant contribution, and at times, can even dominate, the export flux. 
\end{abstract}

@@

Traditionally, POC export is thought to be mainly driven by larger, faster-sinking particles. This reasoning relies on the concept of competing timescales between remineralization and the time required to sink below the base of the mixed layer. The physical processes involved in defining the sinking rate are traditionally assumed to be dominated by gravitational sinking. This is a reasonable assumption if one considers that vertical advective velocities in the ocean are generally much weaker than velocities associated with gravitational sinking.

Recent work demonstrates that the presence of mesoscale eddies can enhance carbon export through two main processes: First, by increasing primary productivity through the injection of nutrients into the euphotic layer, therefore increasing the particle concentration, logically leading to a greater export \citep{Levy_1998}. Second, by generating larger vertical velocities along the front located at the eddy perimeter \citep{vanHaren_2006, Waite_2016}. These studies demonstrate that ocean dynamics potentially play an important role in driving the carbon export from the euphotic layer to the ocean interior. This conclusion is further supported by a recent, yet large, body of literature focusing on ocean dynamics at submeso-scales (1-10 km) and the impact on biogeochemical processes \citep[see reviews in][]{Thomas_2008,Klein_2009,Mahadevan_2016}. Submesoscale dynamics are characterized by Rossby numbers of order 1, indicating that relative vorticity associated with ocean currents exceed the planetary vorticity. These regions typically develop in filaments in areas where sharp density fronts exist \citep{McWilliams_2016}. In this dynamical regime, geostrophic balance fails and a secondary ageostrophic circulation develops across the front, capable of generating large vertical velocities on the order of 100 m/day \citep{FoxKemper_2008,Mahadevan_2016}. On the denser side of the front, the vorticity is cyclonic and associated with downwelling velocities, while anticyclonic vorticity and upwelling is expected on the lighter side of the front. The distribution of relative vorticity across a front is asymmetric and skewed toward cyclonic vorticity \citep{Rudnick_2001}. This asymmetry leads to more localized and more intense downwelling regions, as opposed to weaker and larger upwelling regions \citep{Mahadevan_2006}. Just like at meso-scales, enhanced vertical velocities can generate either a local bloom by supplying nutrients to the sunlit layer of the ocean \citep{Mahadevan_2012}, or can significantly increase the export of POC to the ocean interior \citep{Levy_2012, Estapa_2015,Omand_2015}.

The downwelling velocities $\mathcal{O}$(100 m/day) generated at submeso-scales therefore provide a physical mechanism capable of exporting particles over a larger portion of the particle size spectrum. Through this mechanism, smaller particles can efficiently be exported over timescales shorter than their remineralization timescales, despite their slower sinking velocities. Considering the fact that, although containing less biomass, smaller particles are far more numerous than larger particles, the impact of submesoscale dynamics on the export of POC is potentially significant. Submesoscale processes are, by definition, associated with spatial scales $\mathcal{O}$(1 km) and time scales $\mathcal{O}$(1 day), which makes it difficult to observe. Additionally, and despite recent methodological progress \citep{McDonnell_2010}, direct measurements of particles sinking velocities are difficult to obtain, and are more often than not inferred from key parameters such as particle type, size, and density.

In this study, we rely on a submesoscale-resolving non-hydrostatic model coupled with a particle-tracking module to explore the necessary conditions for slow-sinking particles to significantly contribute to carbon export, as a function of (1) the dynamics of the flow field, (2) the slope of the sinking velocity spectrum, and (3) the remineralization timescale. Particles in the model are prescribed with both a constant and time-varying sinking velocity to mimic a remineralizing behavior. Particles are released in two fundamentally different flow fields in terms of dynamics: one simulation is dominated by mesoscale dynamics, characterized by low Rossby numbers and weak vertical advective velocities, while the other is dominated by submesoscale dynamics, with fine-scale frontal structures and local Rossby numbers $\mathcal{O}$(1). Both simulations and the particle-tracking module are described in Section \ref{sec: Methods}. The impact of particles characteristics and ocean dynamics on the export of POC is quantified  in Section \ref{sec: Results}, and discussed in Section \ref{sec: Discussion}. Section \ref{sec: Conclusions} summarizes the key conclusions of the study.

\section{Methods}
\label{sec: Methods}
\subsection{Model setup and domains}

This study uses a non-hydrostatic, three-dimensional, numerical model called PSOM. Only a brief summary of its characteristics is provided here, as the model is exhaustively described in \cite{Mahadevan_1996a,Mahadevan_1996b}. The model domain is set in a channel configuration with periodic boundaries in the west and east, and solid boundaries in the south and north. The domain covers 112 km in the x-direction, 304 km in the y-direction, and 1000 m in the vertical (Figure \ref{fig: model_domain}). The horizontal resolution is set to 500 m, while the vertical resolution is represented by a stretched grid with 32 levels ranging in thickness from 1.25 m near the surface to 70 m at the lowermost level. The model is integrated numerically in time and evolves the temperature, salinity, free-surface height, pressure, and three-dimensional velocity fields from an initial state, subject to momentum and buoyancy fluxes applied through the surface boundary conditions.

% FIG1
\begin{figure}[ht]
	\centering
	\includegraphics[width = .6\linewidth]{figures/Fig1_model_domain}
	\caption{PSOM model setup. (a) Meridional profile of zonal winds scaling coefficient prescribed to PSOM, with a tapering at the north and south boundaries to reduce upwelling and downwelling, respectively. (b) Restoration factor used to dampen internal waves reflection at boundaries, as well as curl-driven upwelling and downwelling. (c) Surface net heat flux anomaly superimposed to the heat fluxes prescribed at $y = 150$ km.}
	\label{fig: model_domain}
\end{figure}

The zonal component of the  wind stress is prescribed at the surface throughout the model domain ($\tau_y =0$). Winds stress magnitude is, however, tapered at the northern and southern boundaries to avoid excessive Ekman-driven upwelling and downwelling that could contaminate the middle third of the model domain, where the analysis is focused (Figure \ref{fig: model_domain}a). A restoration time scale is prescribed to contain the curl-driven upwelling and downwelling regions generated by the tapering of the wind stress (Figure \ref{fig: model_domain}b). The restored regions also limit the reflection of internal waves at the solid boundaries back into the domain. Net surface heat fluxes are homogeneously prescribed throughout the domain. A linear meridional heat flux anomaly is superimposed to maintain a north/south gradient in net surface heat fluxes  (Figure \ref{fig: model_domain}c). The flux anomaly is zero in the center of the domain and has a slope of 1/24 W/m$^2$/km, estimated from the North American Regional Reanalysis (NARR) product \citep{Mesinger_2006}.

Both wind stress and net heat fluxes are allowed to vary in time. Time series are computed from measurements collected at Station Papa and available through the Pacific Marine Environmental Laboratory\footnote{https://www.pmel.noaa.gov/ocs/Papa}. Climatologies of daily zonal wind stress and net heat fluxes are calculated over the period 2007-2016 to produce a year-long time series. A squared low-pass filter with a cut-off frequency of 8.5 days is applied to both time series to remove high-frequency variability. In all experiments, the simulation is allowed to spin-down, that is with no forcing, for the first 5 days. Forcing is then linearly ramped up between days 5 and 10, to reach realistic values at day 10.% (Figures \ref{fig: Papa_summer} and \ref{fig: Papa_winter}).

Initial background conditions are determined from a three-dimensional gridded field of temperature and salinity from ARGO floats \citep{Gaillard_2009,Gaillard_2016}. ARGO data is averaged monthly over the period 2002-2012 and two different months are used to initialized \textit{Papa\_summer} and \textit{Papa\_winter}. The north-south background density gradient is then intensified into three fronts located at $y = 75$, $y = 150$, and $y = 225$ km (Figure \ref{fig: model_domain}). The amplitude of the density gradient associated with the three fronts are determined from the probability distribution function (PDF) of the density gradients measured by underwater gliders deployed around Station Papa over the period 2008-2010 \citep{Pelland_2016}. To reduce model spin-up time, density fronts are perturbed by a sinusoidal wave with a wavelength $\lambda = 112/n$, where $n$ is the closest integer to the 1st baroclinic deformation radius (n = 2 in this study). Similar PSOM configurations were successfully used in previous studies \citep{Mahadevan_2012,Omand_2015}.

Two main experiments are conducted using the same configuration of PSOM, where only initial conditions and surface forcings are varied: \textit{Papa\_summer} aims at generating ocean dynamics mimicking conditions int the northeast Pacific in the summertime. Summer ocean dynamics are characterized by a flow generally in geostrophic balance, with relatively weak density gradients and low Rossby numbers ($<1$). \textit{Papa\_winter} aims at capturing wintertime ocean conditions in the region. A different dynamical regime is expected to dominate during wintertime, with submesoscale dynamics playing a more important role. Submesoscale dynamics are characterized by sharper density fronts, filament-like features and high localized Rossby number ($>1$) over spatial scales $\mathcal{O}$(1 km). Such dynamics are more likely to be found in the winter time, where mixed layer depth is deeper and density gradients enhanced \citep{Thompson_2016}. The individual characteristics of each of \textit{Papa\_summer} and \textit{Papa\_winter} are detailed below.

@@

In \textit{Papa\_summer}, PSOM is initialized based on climatological ARGO data in April. The magnitude of the density gradient across the front (i.e., frontal sharpness) is set to 3.34$\times$10$^{-6}$ kg/m$^3$/m, which corresponds to the 95$^{th}$ percentile of the PDF of density gradients measured in April from glider data collected in the region (Figure \ref{fig: Papa_summer} and Table \ref{tab: meso_vs_submeso}). The model is run with a timestep of 216 s and is allowed to spin-up for 60 days, allowing for the summer stratification to develop. The model is then run for 30 additional days, saving instantaneous model fields every 3 hours for particle tracking. The month of April is chosen for initialization so the experiment would capture the onset of positive net heat fluxes, and the summer restratification that ensues (Figure \ref{fig: Papa_summer}). In this region, the summer restratification is associated with large primary productivity and therefore large particle production and potentially export \citep[e.g., fecal pallets, dead phytoplankter; ][]{Plant_2016}.

>>>>

%% ------------------------------------------------------------------------ %%
%
%  AUTHORS AND AFFILIATIONS
%
%% ------------------------------------------------------------------------ %%
\authors{M. Dever\affil{1}, D. Nicholson\affil{1}, M. M. Omand\affil{2}, and A. Mahadevan\affil{1}}
\affiliation{1}{Woods Hole Oceanographic Institution, Woods Hole, MA 02543, USA}
\affiliation{2}{Graduate School of Oceanography, University of Rhode Island, Narragansett, Rhode Island, USA}

@@

%% ------------------------------------------------------------------------ %%
%
%  ABSTRACT
%
%% ------------------------------------------------------------------------ %%
\begin{abstract}
% Mat: I think that even with the submesoscale mechanism, export is mainly driven by gravitational sinking. So I'm moving the "Traditionally" to later. 
Export of particulate organic carbon (POC) is  mainly driven by gravitational sinking. Thus, traditionally, it is thought that larger, faster-sinking particles make up most of the POC export flux. However, this need not be the case in a dynamic oceanic flow field, where the ocean velocity can influence the descent rate of particles. Particles with different settling speeds are released in two process-oriented model simulations of an upper ocean eddying flow to evaluate the impact of (1) the particle number size-spectrum slope, and (2) the ocean dynamics on the respective contribution of the different sinking-velocity classes to POC export. The analysis reveals that the leading export mechanism changes from gravitationally-driven to advectively-driven as submesoscale dynamics become more important. The vertical velocity associated with submesoscale dynamics enhances the contribution of slower-sinking particles in POC export. A steeper particle size spectrum, also increases the contribution of smaller, slower-sinking particles. Implementing a remineralization scheme generally decreases the total amount of biomass exported, but its impact is weaker in dynamical regimes where submesoscale dynamics are present and export is advectively-driven. Under specific conditions, remineralization processes counter-intuitively enhance the role of slower-sinking particles to the point where these slower-sinking velocity classes dominate the export, therefore challenging the traditional paradigm for POC export. This study demonstrates that slow-sinking particles are significant contribution, and at times, can even dominate the export flux. 
\end{abstract}

@@

Traditionally, POC export is thought to be mainly driven by larger, faster-sinking, particles capable of gravitationally-sinking below the export depth over timescales shorter than remineralization timescales. This is a reasonable assumption if one considers that vertical advective velocities in the ocean are generally much weaker than velocities associated with gravitational sinking. However, recent work demonstrates that the presence of mesoscale eddies can enhance carbon export through two main processes: First, by increasing primary productivity through the injection of nutrients into the euphotic layer, therefore increasing the particle concentration, logically leading to a greater export \citep{Levy_1998}. Second, by generating larger vertical velocities along the front located at the eddy perimeter \citep{vanHaren_2006, Waite_2016}. These studies demonstrate that ocean dynamics potentially play an important role in driving the carbon export from the euphotic layer to the ocean interior. This conclusion is further supported by a recent, yet large, body of literature focusing on ocean dynamics at submeso-scales (1-10 km) and the impact on biogeochemical processes \citep[see reviews in][]{Thomas_2008,Klein_2009,Mahadevan_2016}. Submesoscale dynamics, characterized by Rossby numbers of order 1, typically develop in filaments in areas where sharp density fronts exist \citep{McWilliams_2016}. In this dynamical regime, geostrophic balance fails and a secondary ageostrophic circulation develops across the front, capable of generating large vertical velocities on the order of 100 m/day \citep{FoxKemper_2008,Mahadevan_2016}. On the denser side of the front, the vorticity is cyclonic and associated with downwelling velocities, while anticyclonic vorticity and upwelling is expected on the lighter side of the front. The distribution of relative vorticity across a front is asymmetric and skewed toward cyclonic vorticity \citep{Rudnick_2001}, leading to more localized and more intense downwelling regions, as opposed to weaker and larger upwelling regions \citep{Mahadevan_2006}. Just like at meso-scales, enhanced vertical velocities can generate either a local bloom by supplying nutrients to the sunlit layer of the ocean \citep{Mahadevan_2012}, or can significantly increase the export of POC to the ocean interior \citep{Levy_2012, Estapa_2015,Omand_2015}.

The downwelling velocities $\mathcal{O}$(100 m/day) generated at submeso-scales provide a physical mechanism capable of competing with gravitational sinking and thus exporting particles over a larger portion of the particle size spectrum. Through this mechanism, smaller particles can efficiently be exported over timescales shorter than their remineralization timescales, despite their slower sinking velocities. Considering the fact that, although containing less biomass, smaller particles are far more numerous than larger particles, the impact of submesoscale dynamics on the export of POC is potentially significant. Submesoscale processes are, by definition, associated with spatial scales $\mathcal{O}$(1 km) and time scales $\mathcal{O}$(1 day), which makes it difficult to observe in-situ. Additionally, and despite recent methodological progress \citep{McDonnell_2010}, direct measurements of particles sinking velocities are difficult to obtain, and are more often than not inferred from key parameters such as particle type, size, and density.

In this study, we therefore rely on a submesoscale-resolving non-hydrostatic model coupled with a particle-tracking module to explore the necessary conditions for slow-sinking particles to significantly contribute to carbon export, as a function of (1) the dynamics of the flow field, (2) the slope of the sinking velocity spectrum, and (3) the remineralization timescale. Particles in the model are prescribed with both a constant and time-varying sinking velocity to mimic a remineralizing behavior. Particles are released in two fundamentally different flow fields in terms of dynamics based on observed conditions in the Northeast Pacific in the summer, where ocean dynamics are characterized by low Rossby numbers and weak vertical advective velocities, and in the winter, where ocean dynamics include submesoscale frontal structures and local Rossby numbers $\mathcal{O}$(1). Both simulations and the particle-tracking module are described in Section \ref{sec: Methods}. The impact of particles characteristics and ocean dynamics on the export of POC is quantified in Section \ref{sec: Results}, and discussed in Section \ref{sec: Discussion}. Section \ref{sec: Conclusions} summarizes the key conclusions of the study.

\section{Methods}
\label{sec: Methods}
\subsection{Model setup and domain}

This study uses a non-hydrostatic, three-dimensional, numerical model called PSOM \citep{Mahadevan_1996a,Mahadevan_1996b}. To simulate an eddy field that is representative of the Northeast Pacific Ocean, the model domain is set in a channel configuration with periodic east-west boundaries, and solid boundaries in the south and north. The domain covers 112 km in the x-direction, 304 km in the y-direction, and 1000 m in the vertical (Figure \ref{fig: model_domain}). The horizontal resolution is 500 m, while a stretched grid is used in the vertical with 32 levels ranging in thickness from 1.25 m near the surface to 70 m at the lowermost level. The model is integrated numerically in time and evolves the temperature, salinity, free-surface height, pressure, and three-dimensional velocity field from an initial state, subject to momentum and buoyancy fluxes applied through the surface boundary conditions.

% FIG1
\begin{figure}[ht]
	\centering
	\includegraphics[width = .6\linewidth]{figures/Fig1_model_domain}
	\caption{PSOM model setup. (a) Meridional profile of zonal wind scaling coefficient that multiplies $\tau_x$, with a tapering at the north and south boundaries to reduce upwelling and downwelling, respectively. (b) Restoration factor (color shading) used to dampen internal wave reflection at boundaries as well as curl-driven upwelling and downwelling, along with constant density contours showing the three fronts (black line). (c) Surface net heat flux anomaly superimposed to the heat fluxes prescribed at $y = 150$ km.}
	\label{fig: model_domain}
\end{figure}

Time-varying wind stress and heat flux are prescribed at the surface boundary. Time series are computed from measurements collected at Station Papa and available through the Pacific Marine Environmental Laboratory\footnote{https://www.pmel.noaa.gov/ocs/Papa}. Climatologies of daily wind stress and net heat fluxes are calculated over the period 2007-2016 to produce a year-long time series. A squared low-pass filter with a cut-off frequency of 8.5 days is applied to both time series to remove high-frequency variability. In all numerical experiments, the simulations are first allowed to spin-down for the first 5 days with no forcing apply at the surface boundary. Surface wind stress and heat fluxes are then linearly ramped up between days 5 and 10 of the simulation, to reach realistic values at day 10. % (Figures \ref{fig: Papa_summer} and \ref{fig: Papa_winter}).
While the meriodional component, $\tau_y$, is set to zero, the zonal component of the wind stress, $\tau_x$, is prescribed at the surface throughout the model domain and is tapered at the northern and southern boundaries to avoid excessive Ekman-driven upwelling and downwelling that could contaminate the middle third of the model domain, where the analysis is focused (Figure \ref{fig: model_domain}a). A restoration time scale is prescribed to contain the curl-driven upwelling and downwelling regions generated by the tapering of the wind stress, as well as to limit internal wave reflection at the solid boundaries back into the domain (Figure \ref{fig: model_domain}b). Net surface heat fluxes are homogeneously prescribed throughout the domain, over which a linear meridional heat flux anomaly is superimposed to maintain a north-south gradient (Figure \ref{fig: model_domain}c). The flux anomaly is zero in the center of the domain, associated with the latitude of Station Papa, and has a meridional slope of 1/24 W/m$^2$/km, which was determined from the North American Regional Reanalysis (NARR) product \citep{Mesinger_2006}.

Initial hydrographic conditions are determined from a three-dimensional gridded field of temperature and salinity from ARGO floats \citep{Gaillard_2009,Gaillard_2016}. ARGO data is averaged monthly over the period 2002-2012 and two different months are used to initialize the two main numerical experiments for this study: Climatological conditions in April are used to initialize the \textit{Papa\_summer} experiment, while January climatological conditions are used to initialize the \textit{Papa\_winter} experiment (Table \ref{tab: meso_vs_submeso}). The north-south background density gradient is then intensified into three fronts located at $y = 75$, $y = 150$, and $y = 225$ km (Figure \ref{fig: model_domain}). The amplitude of the density gradient associated with the three fronts is determined from the probability distribution function (PDF) of the density gradients measured by underwater gliders deployed around Station Papa over the period 2008-2010 \citep{Pelland_2016}. To reduce model spin-up time, density fronts are perturbed by a sinusoidal wave with a wavelength close to the 1st baroclinic deformation radius ($\lambda = $ 66 km). Similar PSOM configurations were successfully used in previous studies \citep{Mahadevan_2012,Omand_2015}.

Two main experiments are conducted using the same configuration of PSOM, where only initial conditions and surface forcings are varied: \textit{Papa\_summer} aims at generating ocean dynamics mimicking conditions int the northeast Pacific in the summertime. Summer ocean dynamics are characterized by a flow generally in geostrophic balance, with relatively weak density gradients and low Rossby numbers ($\ll1$). \textit{Papa\_winter} aims at capturing wintertime ocean conditions in the region. A different dynamical regime is expected to dominate during wintertime, with submesoscale dynamics playing a more important role. Submesoscale dynamics are characterized by sharper density fronts, filament-like features and localized Rossby number $\mathcal{O}$(1) over spatial scales $\mathcal{O}$(1 km). Such dynamics are more likely to be found in the winter time, where mixed layers are deeper and lateral density gradients enhanced \citep{Thompson_2016}. The individual characteristics of each of \textit{Papa\_summer} and \textit{Papa\_winter} are detailed below.

@@

In \textit{Papa\_summer}, PSOM is initialized based on climatological ARGO data in April. The magnitude of the density gradient across the front (i.e., frontal sharpness) is set to 3.34$\times$10$^{-6}$ kg/m$^3$/m, which corresponds to the 95$^{th}$ percentile of the PDF of density gradients measured in April from glider data collected in the region (Figure \ref{fig: Papa_summer} and Table \ref{tab: meso_vs_submeso}). The model is run with a timestep of 216 s and is allowed to spin-up for 60 days, allowing summer stratification to develop. The model is then run for 30 additional days, saving instantaneous model fields every 3 hours for particle tracking. The month of April is chosen for initialization so the experiment would capture the onset of positive net heat fluxes, and the summer restratification that ensues in July-August (Figure \ref{fig: Papa_summer}). In this region, the summer stratification is associated with large primary productivity and therefore large particle production and potentially export \citep[e.g., fecal pallets, dead phytoplankter; ][]{Plant_2016}.
